Exercising choice over feedback schedules during practice is not advantageous for motor learning

### 2022-08-14 docs/revision-1/r1_manuscript.tex

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Self-controlled, Knowledge of results, OPTIMAL theory, Retention\newline\indent Word count: 2978 for Introduction, Results, and Discussion}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{pdflscape}
\usepackage{setspace}
\pagewiselinenumbers
\raggedbottom
\renewcommand\author[1]{}
\renewcommand\affiliation[1]{}
\authorsnames[1, 1, 1, 1, 1, 1, {2,3,4,5}, 1]{Laura St. Germain, Brad McKay, Andrew Poskus, Allison Williams, Olena Leshchyshen, Sherry Feldman, Joshua G.A. Cashaback, Michael J. Carter\vspace{2ex}}
\authorsaffiliations{{Department of Kinesiology, McMaster University}, {Department of Biomedical Engineering, University of Delaware}, {Department of Mechanical Engineering, University of Delaware}, {Biomechanics and Movements Science Program, University of Delaware}, {Interdisciplinary Neuroscience Graduate Program, University of Delaware}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Exercising choice over feedback schedules during practice is not advantageous for motor learning},
  pdflang={en-EN},
  pdfkeywords={Self-controlled, Knowledge of results, OPTIMAL theory, Retention},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\noindent \textbf{Corresponding authors:} Michael J. Carter (\href{mailto:michaelcarter@mcmaster.ca}{\nolinkurl{michaelcarter@mcmaster.ca}}) or Laura St.~Germain (\href{mailto:stgerml@mcmaster.ca}{\nolinkurl{stgerml@mcmaster.ca}})

@@

\abstract{%
The idea that there is a self-controlled learning advantage, where individuals demonstrate improved motor learning after exercising choice over an aspect of practice compared to no-choice groups, has different causal explanations according to the OPTIMAL theory or an information-processing perspective. Within OPTIMAL theory, giving learners choice is considered an autonomy-supportive manipulation that enhances expectations for success and intrinsic motivation. In the information-processing view, choice allows learners to engage in performance-dependent strategies that reduce uncertainty about task outcomes. To disentangle these potential explanations, we provided participants in choice and yoked groups with error or graded feedback (Experiment 1) and binary feedback (Experiment 2) while learning a rapid reaching task with spatial and timing goals. Across both experiments (N = 228 participants), we did not find evidence to support a self-controlled learning advantage. Exercising choice during practice did not increase perceptions of autonomy, competence, or intrinsic motivation, nor did it lead to more accurate error estimation skills. Both error and graded feedback facilitated skill acquisition and learning, whereas no improvements from pre-test performance were found with binary feedback. Finally, the impact of graded and binary feedback on perceived competence highlights a potential dissociation of motivational and informational roles of feedback. Although our results regarding self-controlled practice conditions are difficult to reconcile with either the OPTIMAL theory or the information-processing perspective, they are consistent with a growing body of evidence that strongly suggests self-controlled conditions are not an effective approach to enhance motor performance and learning.
}

@@

The underlying source of errors in skilled actions are often ambiguous and difficult to assign as the learner must rely on noisy and delayed sensory information. Feedback from an external source, such as a coach or computer display, can facilitate or augment this process (Sigrist et al., 2013). Knowledge of results feedback (Salmoni et al., 1984) can provide varying amounts of information to learners depending on its characteristics. Error feedback provides precise information about the magnitude and direction of the error (e.g., -42 cm), graded feedback provides coarse information about either the magnitude or direction of the error (e.g., ``too far''), and binary feedback indicates only success or failure information (e.g., ``miss'') (Luft, 2014). When to provide this feedback is often decided by an external agent; however, this feedback decision can also be made by the learner, a form of self-controlled learning. These self-controlled feedback schedules have typically enhanced motor skill learning compared to yoked feedback schedules, wherein learners experience the feedback schedule created by a self-controlled counterpart, but without any choice (see Sanli et al., 2013; Ste-Marie et al., 2020 for reviews).

Why self-controlled learning advantages emerge has garnered considerable attention in the motor skill learning literature. Within their OPTIMAL (Optimizing performance through intrinsic motivation and attentional learning) theory of motor learning, Wulf and Lewthwaite (2016) have argued that providing participants the opportunity to exercise choice, as in a self-controlled group, creates a virtuous cycle. Specifically, choice leads to increased (perceived) autonomy, leading to enhanced expectancies (e.g., perceived competence) and increased (intrinsic) motivation. These motivational influences lead to improved motor performance, creating a positive feedback loop that ultimately enhances motor learning compared to those not given the same choice opportunities. Support for this view has been drawn from experimental work where participants exercise choice over task-irrelevant or incidental choices. Exercising choice over the color of golf balls to putt (Lewthwaite et al., 2015 Experiment 1) or the mat underneath a target (Wulf et al., 2018 Experiment 1), which picture to hang in a lab (Lewthwaite et al., 2015 Experiment 2), hand order in a maximal force production task (Iwatsuki et al., 2017), which photos to look at while running (Iwatsuki et al., 2018), and the order of exercises to perform (Wulf et al., 2014) have been suggested to improve motor performance or learning. Other research, however, have failed to replicate this benefit of task-irrelevant or incidental choices on motor performance or learning (Carter \& Ste-Marie, 2017a; Grand et al., 2017; McKay \& Ste-Marie, 2020a, 2020b).

@@

To dissociate between the motivational and information-processing accounts of the self-controlled learning advantage, we provided participants in choice and yoked (i.e., no-choice) groups with error or graded feedback (Experiment 1) and binary feedback (Experiment 2) during the acquisition of a rapid aiming task. Motor learning was assessed using delayed (\textasciitilde24 hours) retention and transfer tests. If the OPTIMAL theory is correct, we hypothesized that the characteristics of one's feedback schedule \emph{would not matter} for the self-controlled learning advantage as this advantage arises from the opportunity for choice--a common feature of all choice groups. Thus, we predicted all choice groups would demonstrate superior performance and learning compared to the yoked groups. Alternatively, if the information-processing account is correct, we hypothesized that the characteristics of one's feedback schedule \emph{would matter} for the self-controlled learning advantage as feedback with greater informational value would be more effective for reducing uncertainties about movement outcomes. Thus, we predicted that choice over an error feedback schedule would be the most effective pairing for performance and learning. We also included self-report measures of perceptions of autonomy, competence, and intrinsic motivation, and assessments of error estimation abilities to respectively test auxiliary assumptions of the OPTIMAL theory and information-processing explanations.

@@

One hundred and fifty-two right-handed (Oldfield, 1971), healthy adults participated in Experiment 1 (\(M_{age}\) = 20.64 years, \(SD_{age}\) = 2.45, 88 females). Sample size was determined from an \emph{a-priori} power analysis using the \emph{ANOVA: fixed effects, main effects and interactions} option in G*Power (Faul et al., 2009) with the following parameters: \(\alpha\) = 0.05, \(\beta\) = .20, \(f\) = 0.23, numerator = 1, and groups = 2. This revealed a required sample of 151 participants. Our chosen effect size was based on a meta-analytic estimate (\(f\) = .32) by McKay et al. (2014); however, we used a more conservative estimate given the uncertainty of how choice would interact with our feedback characteristic manipulation. Participants were compensated \$15 CAD or with course-credit for their time. All participants gave written informed consent and the experiment was approved by McMaster University's Research Ethics Board.

@@

Data collection consisted of two sessions separated by approximately 24 hours.\footnote{Six participants (three Choice+Error-Feedback and three Choice+Graded-Feedback) had their second session completed approximately 48 hours later because a snowstorm closed the University.} Session one included a pre-test (12 trials) and an acquisition phase (72 trials). Session two included the delayed retention (12 trials) and transfer (12 trials) tests. No feedback about motor performance was provided in pre-test, retention, or transfer. Prior to the pre-test, all participants received instructions about the task and its associated spatial and timing goals. Additionally, half of the participants in each group were randomly selected to estimate their performance on the spatial and timing goals after each trial in the pre-test. Only a subset of participants were asked to estimate their performance in pre-test to mitigate the potential that doing so would prompt participants to adopt this strategy during the experiment as error estimation has been suggested (e.g., Chiviacowsky \& Wulf, 2005) to be adopted spontaneously by participants controlling their feedback schedule. However, asking participants to estimate their performance during pre-test is necessary to be able to assess how this skill develops as a function of one's practice condition.

@@

A typical acquisition trial (see Figure \ref{fig:fig1}) began with the current trial number displayed (500 ms), followed by a visual ``Get Ready!'' and a visual go-signal (800 ms apart). Participants were free to begin their movement when ready following the visual go-signal (i.e., green circle) as this was not a reaction time task. The computer screen was blank while participants made their movement. When participants returned to the starting position, a red circle was displayed on the monitor. Following a 2000 ms feedback delay interval, the feedback decision prompt was presented for the self-controlled groups. If feedback was not selected, a blank screen was displayed for 3000 ms. If feedback was selected (or imposed on the yoked groups), it was also displayed for 3000 ms.

Before the retention and transfer tests, participants were reminded about the task and its associated goals. All participants were asked to estimate their performance after each trial in retention and transfer. After the pre-test, trials 12 and 72 in acquisition, and before the delayed retention test, participants answered a series of questions pertaining to perceived competence, task interest and enjoyment, and perceived autonomy. The perceived competence and task interest and enjoyment questions were from the Intrinsic Motivation Inventory (McAuley et al., 1989; Ryan, 1982) and the perceived autonomy questions were used in earlier work (Barros et al., 2019; Carter \& Ste-Marie, 2017a). Cronbach's alpha values for each questionnaire at each time point are reported in Table \ref{tab:table1}.

@@

\caption{\small \onehalfspacing \textbf{Overview of a typical acquisition trial for the choice groups.} The sequence of events a participant in the choice groups experienced during the acquisition phase. Trials began by informing participants the trial number (500 ms) they were on in acquisition. Shortly after, the text ``Get Ready!'' appeared on the screen and 800 ms later a visual go-signal was presented in the form of a green circle in the center of the screen. Participants began their movement when ready after seeing the visual go-signal as we were not interested in reaction time. While participants completed their rapid out-and-back movement, the computer screen was blank. Upon returning to the starting position, a red circle appeared in the center of the screen. A 2000 ms feedback delay interval was used and this interval was followed by the feedback prompt. The feedback prompt also displayed an updated counter representing the number of feedback trials they had left. If the number of remaining feedback trials matched the number of acquisition trials left, these trials automatically defaulted to feedback trials. On trials where feedback was not requested, a blank screen \textbf{(A)} was shown for 3000 ms. When feedback was selected, feedback was provided for both the spatial and timing goals according to their experimental group. The error feedback group \textbf{(B)} saw their constant error, the graded feedback group \textbf{(C)} saw either ``too far'' or ``too short'' for the spatial goal and ``too fast'' or ``too slow'' for the timing goal, and the binary feedback group \textbf{(D)} saw either ``hit'' or ``miss'' for the task goals. The sequence of events was the same for the yoked groups with the exception they did not see a feedback prompt. The sequence of events was similar in pre-test, retention, and transfer except all trials were no-feedback trials.}\label{fig:fig1}
\end{figure}

@@

@@

To test for performance differences in pre-test, retention, and transfer, total error for the spatial and timing goals were analyzed in separate mixed ANOVAs (\emph{Experiment 1:} 2 Choice x 2 Feedback x 3 Test; \emph{Experiment 2:} 2 Choice x 3 Test). To test for performance differences during acquisition, total error for the spatial and timing goals during acquisition were analyzed in separate mixed ANOVAs (\emph{Experiment 1:} 2 Choice x 2 Feedback x 6 Block; \emph{Experiment 2:} 2 Choice x 6 Block). Our primary psychological outcome variables were intrinsic motivation (i.e., interest/enjoyment), perceived competence, and perceived autonomy. The mean score of the responses for these constructs at each time point was calculated for each participant and analyzed in separate mixed ANOVAs (\emph{Experiment 1:} 2 Choice x 2 Feedback x 4 Time; \emph{Experiment 2:} 2 Choice x 4 Time). Of secondary interest, error estimation abilities were assessed as total error between a participant's estimation and actual performance in pre-test (50\% of the participants in each group in Experiments 1 and 2), retention, and transfer (see \textbf{Supplementary A}).

@@

Model diagnostics of total error for the spatial and timing goals revealed skewed distributions. We therefore conducted sensitivity analyses using the shift function, which is a robust statistical method well-suited for skewed distributions (Rousselet \& Wilcox, 2020; Wilcox, 2021). The results of these analyses (see \textbf{Supplementary B}) were consistent with those of the mixed ANOVAs, which we report below.

@@

Spatial (Fig. \ref{fig:fig2}A) and timing (Fig. \ref{fig:fig2}B) error decreased from the pre-test to the retention and transfer tests. There was a main effect of Test for spatial error, \(F(1.33,196.52) = 40.20\), \(p < .001\), \(\eta_{G}^2 = .138\), where performance was less errorful in retention and transfer than pre-test (\(p\)'s \textless{} .001) and performance in retention was better than transfer (\(p\) \textless{} .001). A main effect of Test was also found for timing error, \(F(1.08,160.23) = 81.21\), \(p < .001\), \(\eta_{G}^2 = .245\), with pre-test performance more errorful than both retention and transfer (\(p\)'s \textless{} .001), and retention was less errorful than transfer (\(p\) \textless{} .001).

@@

Spatial (Fig. \ref{fig:fig3}A) and timing (Fig. \ref{fig:fig3}B) error did not change considerably from the pre-test to the retention and transfer tests. There were no significant main effects or interactions for either task goals.

All groups of participants improved their performance of the spatial goal during the acquisition phase (Fig. \ref{fig:fig2}A). This was supported by a significant main effect of Block, \(F(2.41,357.13) = 60.18\), \(p < .001\), \(\eta_{G}^2 = .130\), where block 1 was less accurate than all other blocks (\(p\)'s \(< .001\)), block 2 was less accurate than all subsequent blocks (\(p\)'s \(\leq .021\)), and blocks 3 and 4 were more errorful than block 6 (\(p\)'s \(\leq .015\)). Timing error also decreased during the acquisition period (Fig. \ref{fig:fig2}B). The significant main effect of Block, \(F(1.75,259.59) = 55.44\), \(p < .001\), \(\eta_{G}^2 = .138\), was superseded by a significant Feedback x Block interaction, \(F(1.75, 259.59) = 3.56\), \(p = .035\), \(\eta_{G}^2 = .010\). Post hoc comparisons showed that timing error for those receiving error feedback was reduced from block 1 in all subsequent blocks (\(p\)'s \(< .001\)), but performance plateaued from block 2 onward in acquisition (\(p\)'s \(\geq .257\)). Timing error for the participants that received graded feedback was also reduced from block 1 in all subsequent blocks (\(p\)'s \(< .001\)); however, these participants continued to improve across acquisition blocks as block 2 was more errorful than blocks 3 to 6 (\(p\)'s \(\leq .028\)).

@@

\caption{\small \onehalfspacing \textbf{Experiment 1 data.} The Choice with error feedback (Choice+Error) group is shown in dark blue circles, the Choice with graded feedback (Choice+Graded) group is shown in light blue squares, the Yoked with error feedback (Yoked+Error) group is shown in red triangles, and the Yoked with graded feedback (Yoked+Graded) group is shown in yellow crosses. Error bars denote 95\% bootstrapped confidence intervals. \textbf{(A)} Spatial total error (degrees) and \textbf{(B)} timing total error (ms) averaged across blocks and participants within each group. Dotted vertical lines denote the different experimental phases. Pre-test and acquisition occurred on Day 1 and retention and transfer occurred approximately 24-hours later on Day 2. Self-reported scores for perceived autonomy \textbf{(C)}, perceived competence \textbf{(D)}, and intrinsic motivation \textbf{(E)} after the pre-test and after blocks 1 and 6 of acquisition on Day 1, and before the retention test on Day 2. Scores could range on a Likert scale from 1 (Strongly disagree) to 7 (Strongly agree). Dots represent individual data points.}\label{fig:fig2}
\end{figure}

Spatial (Fig. \ref{fig:fig3}A) and timing (Fig. \ref{fig:fig3}B) error remained relatively flat from block 1 to block 6 in the acquisition period. None of the main effects or interactions for either task goals were significant.

@@

\caption{\small \onehalfspacing \textbf{Experiment 2 data.} The Choice with binary feedback (Choice+Binary) group is shown in green circles and the Yoked with binary feedback (Yoked+Binary) group is shown in purple squares. Error bars denote 95\% bootstrapped confidence intervals. \textbf{(A)} Spatial total error (degrees) and \textbf{(B)} timing total error (ms) averaged across blocks and participants within each group. Dotted vertical lines denote the different experimental phases. Pre-test and acquisition occurred on Day 1 and retention and transfer occurred approximately 24-hours later on Day 2. Self-reported scores for perceived autonomy \textbf{(C)}, perceived competence \textbf{(D)}, and intrinsic motivation \textbf{(E)} after the pre-test and after blocks 1 and 6 of acquisition on Day 1, and before the retention test on Day 2. Scores could range on a Likert scale from 1 (Strongly disagree) to 7 (Strongly agree). Dots represent individual data points.}\label{fig:fig3}
\end{figure}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Self-controlled, Knowledge of results, OPTIMAL theory, Retention}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{pdflscape}
\usepackage{setspace}
\raggedbottom
\renewcommand\author[1]{}
\renewcommand\affiliation[1]{}
\authorsnames[1, 1, 1, 1, 1, 1, {2,3,4,5}, 1]{Laura St. Germain, Brad McKay, Andrew Poskus, Allison Williams, Olena Leshchyshen, Sherry Feldman, Joshua G.A. Cashaback, Michael J. Carter\vspace{2ex}}
\authorsaffiliations{{Department of Kinesiology, McMaster University}, {Department of Biomedical Engineering, University of Delaware}, {Department of Mechanical Engineering, University of Delaware}, {Biomechanics and Movements Science Program, University of Delaware}, {Interdisciplinary Neuroscience Graduate Program, University of Delaware}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Exercising choice over feedback schedules during practice is not advantageous for motor learning},
  pdflang={en-EN},
  pdfkeywords={Self-controlled, Knowledge of results, OPTIMAL theory, Retention},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\noindent \textbf{Corresponding authors:} Laura St.~Germain (\href{mailto:stgerml@mcmaster.ca}{\nolinkurl{stgerml@mcmaster.ca}}) or Michael J. Carter (\href{mailto:cartem11@mcmaster}{\nolinkurl{cartem11@mcmaster}}; \href{mailto:motorlab@mcmaster.ca}{\nolinkurl{motorlab@mcmaster.ca}})

@@

\abstract{%
The idea that there is a self-controlled learning advantage, where individuals demonstrate improved motor learning after exercising choice over an aspect of practice compared to no-choice groups, has different causal explanations according to the OPTIMAL theory or an information-processing perspective. Within OPTIMAL theory, giving learners choice is considered an autonomy-supportive manipulation that enhances expectations for success and intrinsic motivation. In the information-processing view, choice allows learners to engage in performance-dependent strategies that reduce uncertainty about task outcomes. To disentangle these potential explanations, we provided participants in choice and yoked groups with error or graded feedback (Experiment 1) and binary feedback (Experiment 2) while learning a \textcolor{blue}{novel motor} task with spatial and timing goals. Across both experiments (N = 228 participants), we did not find evidence to support a self-controlled learning advantage. Exercising choice during practice did not increase perceptions of autonomy, competence, or intrinsic motivation, nor did it lead to more accurate error estimation skills. Both error and graded feedback facilitated skill acquisition and learning, whereas no improvements from pre-test performance were found with binary feedback. Finally, the impact of graded and binary feedback on perceived competence highlights a potential dissociation of motivational and informational roles of feedback. Although our results regarding self-controlled practice conditions are difficult to reconcile with either the OPTIMAL theory or the information-processing perspective, they are consistent with a growing body of evidence that strongly suggests self-controlled conditions are not an effective approach to enhance motor performance and learning.
}

@@

The underlying source of errors in skilled actions are often ambiguous and difficult to assign as the learner must rely on noisy and delayed sensory information. Feedback from an external source, such as a coach or computer display, can facilitate or augment this process (Sigrist et al., 2013). Knowledge of results feedback (Salmoni et al., 1984) can provide varying amounts of information to learners depending on its characteristics. Error feedback provides precise information about the magnitude and direction of the error (e.g., -42 cm), graded feedback provides coarse information about either the magnitude or direction of the error (e.g., ``too far''), and binary feedback indicates only success or failure information (e.g., ``miss'') (Luft, 2014).\footnote{\textcolor{blue}{Others have referred to error feedback as quantitative feedback and graded feedback as qualitative feedback} (e.g., Magill \& Wood, 1986)\textcolor{blue}{. We use the terminology error, graded, and binary feedback because graded and binary feedback are different forms of qualitative feedback.}} When to provide this feedback is often decided by an external agent; however, this feedback decision can also be made by the learner, a form of self-controlled learning. These self-controlled feedback schedules have typically enhanced motor skill learning compared to yoked feedback schedules, wherein learners experience the feedback schedule created by a self-controlled counterpart, but without any choice (see Sanli et al., 2013; Ste-Marie et al., 2020 for reviews).

Why self-controlled learning advantages emerge has garnered considerable attention in the motor skill learning literature. Within their OPTIMAL (Optimizing performance through intrinsic motivation and attentional learning) theory of motor learning, Wulf and Lewthwaite (2016) have argued that providing participants the opportunity to exercise choice, as in a self-controlled group, creates a virtuous cycle. Specifically, choice leads to increased (perceived) autonomy, leading to enhanced expectancies (e.g., perceived competence) and increased (intrinsic) motivation. These motivational influences lead to improved motor performance, creating a positive feedback loop that ultimately enhances motor learning compared to those not given the same choice opportunities. Support for this view has been drawn from experimental work where participants exercise choice over task-irrelevant or incidental choices. Exercising choice over the color of golf balls to putt (Lewthwaite et al., 2015 Experiment 1) or the mat underneath a target (Wulf et al., 2018 Experiment 1), which picture to hang in a lab (Lewthwaite et al., 2015 Experiment 2), hand order in a maximal force production task (Iwatsuki et al., 2017), which photos to look at while running (Iwatsuki et al., 2018), and the order of exercises to perform (Wulf et al., 2014) have been suggested to improve motor performance or learning. Other research, however, have failed to replicate this benefit of task-irrelevant or incidental choices on motor performance or learning (Carter \& Ste-Marie, 2017a; Grand et al., 2017; McKay \& Ste-Marie, 2020, 2022).

@@

To dissociate between the motivational and information-processing accounts of the self-controlled learning advantage, \textcolor{blue}{we manipulated the amount of information participants in choice and yoked (i.e., no-choice) groups experienced with their feedback schedule during acqusition of a novel motor task. In Experiment 1, participants received error or graded feedback to assess how high and moderate levels of informational value impact the self-controlled learning advantage. Given both error and graded feedback provide salient information about how to correct one's behavior relative to the task goal (i.e., both generate an error signal), in Experiment 2 we provided participants with binary feedback. As binary feedback is devoid of information about the necessary change to improve one's behavior (i.e., does not generate an error signal), we could better isolate the motivational nature of choice to test between the two explanations for the self-controlled learning advantage.} Motor learning was assessed using delayed (\textasciitilde24 hours) retention and transfer tests. If the OPTIMAL theory is correct, we hypothesized that the characteristics of one's feedback schedule \emph{would not matter} for the self-controlled learning advantage as this advantage arises from the opportunity for choice--a common feature of all choice groups. Thus, we predicted all choice groups would demonstrate superior performance and learning compared to the yoked groups. Alternatively, if the information-processing account is correct, we hypothesized that the characteristics of one's feedback schedule \emph{would matter} for the self-controlled learning advantage as feedback with greater informational value would be more effective for reducing uncertainties about movement outcomes. Thus, we predicted that choice over an error feedback schedule would be the most effective pairing for performance and learning. We also included self-report measures of perceptions of autonomy, competence, and intrinsic motivation, and assessments of error estimation abilities to respectively test auxiliary assumptions of the OPTIMAL theory and information-processing explanations.

@@

One hundred and fifty-two right-handed (Oldfield, 1971), healthy adults participated in Experiment 1 (\(M_{age}\) = 20.64 years, \(SD_{age}\) = 2.45, 88 females). Sample size was determined from an \emph{a-priori} power analysis using the \emph{ANOVA: fixed effects, main effects and interactions} option in G*Power (Faul et al., 2009) with the following parameters: \(\alpha\) = 0.05, \(\beta\) = .20, \(f\) = 0.23, numerator = 1, and groups = \textcolor{blue}{4}. This revealed a required sample of 151 participants. Our chosen effect size was based on a meta-analytic estimate (\emph{f} = .32) by McKay et al. (2014); however, we used a more conservative estimate given the uncertainty of how choice would interact with our feedback characteristic manipulation. Participants were compensated \$15 CAD or with course-credit for their time. All participants gave written informed consent and the experiment was approved by McMaster University's Research Ethics Board.

@@

Data collection consisted of two sessions separated by approximately 24 hours.\footnote{Six participants (three Choice+Error-Feedback and three Choice+Graded-Feedback) had their second session completed approximately 48 hours later because a snowstorm closed the University.} Session one included a pre-test (12 trials) and an acquisition phase (72 trials). Session two included the delayed retention (12 trials) and transfer (12 trials) tests. No feedback about motor performance was provided in pre-test, retention, or transfer. Prior to the pre-test, all participants received instructions about the task and its associated spatial and timing goals. Additionally, half of the participants in each group were randomly selected to \textcolor{blue}{verbally} estimate their performance on the spatial and timing goals after each trial in the pre-test. Only a subset of participants were asked to estimate their performance in pre-test to mitigate the potential that doing so would prompt participants to adopt this strategy during the experiment as error estimation has been suggested (e.g., Chiviacowsky \& Wulf, 2005) to be adopted spontaneously by participants controlling their feedback schedule. However, asking participants to estimate their performance during pre-test is necessary to be able to assess how this skill develops as a function of one's practice condition.

@@

A typical acquisition trial (see Figure \ref{fig:fig1}) began with the current trial number displayed (500 ms), followed by a visual ``Get Ready!'' and a visual go-signal (800 ms apart). Participants were free to begin their movement when ready following the visual go-signal (i.e., green circle) as this was not a reaction time task. The computer screen was blank while participants made their movement. When participants returned to the starting position, a red circle was displayed on the monitor. Following a 2000 ms feedback delay interval, the feedback decision prompt was presented for the self-controlled groups. \textcolor{blue}{The number of remaining feedback trials was also displayed during this feedback delay interval.} If feedback was not selected, a blank screen was displayed for 3000 ms. If feedback was selected via \textcolor{blue}{verbal response} (or imposed on the yoked groups), it was also displayed for 3000 ms.

\clearpage

@@

\caption{\small \normalfont \onehalfspacing \textbf{Overview of a typical acquisition trial for the choice groups.} The sequence of events a participant in the choice groups experienced during the acquisition phase. Trials began by informing participants the trial number (500 ms) they were on in acquisition. Shortly after, the text ``Get Ready!'' appeared on the screen and 800 ms later a visual go-signal was presented in the form of a green circle in the center of the screen. Participants began their movement when ready after seeing the visual go-signal as we were not interested in reaction time. While participants completed their rapid out-and-back movement, the computer screen was blank. Upon returning to the starting position, a red circle appeared in the center of the screen. A 2000 ms feedback delay interval was used and this interval was followed by the feedback prompt. The feedback prompt also displayed an updated counter representing the number of feedback trials they had left. If the number of remaining feedback trials matched the number of acquisition trials left, these trials automatically defaulted to feedback trials. On trials where feedback was not requested, a blank screen \textbf{(A)} was shown for 3000 ms. When feedback was selected via \textcolor{blue}{verbal response}, feedback was provided for both the spatial and timing goals according to their experimental group. The error feedback group \textbf{(B)} saw their constant error, the graded feedback group \textbf{(C)} saw either ``too far'' or ``too short'' for the spatial goal and ``too fast'' or ``too slow'' for the timing goal, and the binary feedback group \textbf{(D)} saw either ``hit'' or ``miss'' for the task goals. The sequence of events was the same for the yoked groups with the exception they did not see a feedback prompt. The sequence of events was similar in pre-test, retention, and transfer except all trials were no-feedback trials.}\label{fig:fig1}
\end{figure}

\clearpage

@@

Before the retention and transfer tests, participants were reminded about the task and its associated goals. All participants were asked to \textcolor{blue}{verbally} estimate their performance after each trial in retention and transfer. After the pre-test, trials 12 and 72 in acquisition, and before the delayed retention test, participants \textcolor{blue}{verbally} answered a series of questions pertaining to perceived competence, task interest and enjoyment, and perceived autonomy.\footnote{\textcolor{blue}{The questionnaires can be found in the publicly available project repository in the materials directory.}} The perceived competence and task interest and enjoyment questions were from the Intrinsic Motivation Inventory (McAuley et al., 1989; Ryan, 1982) and the perceived autonomy questions were used in earlier work (Barros et al., 2019; Carter \& Ste-Marie, 2017a; St. Germain et al., 2022). Cronbach's alpha values for each questionnaire at each time point are reported in Table \ref{tab:table1}.

@@

To test for performance differences in pre-test, retention, and transfer, total error for the spatial and timing goals were analyzed in separate mixed ANOVAs (\emph{Experiment 1:} 2 Choice x 2 Feedback x 3 Test; \emph{Experiment 2:} 2 Choice x 3 Test). To test for performance differences during acquisition, total error for the spatial and timing goals during acquisition were analyzed in separate mixed ANOVAs (\emph{Experiment 1:} 2 Choice x 2 Feedback x 6 Block; \emph{Experiment 2:} 2 Choice x 6 Block). Model diagnostics of total error for the spatial and timing goals revealed skewed distributions. We therefore conducted sensitivity analyses using the shift function, which is a robust statistical method well-suited for skewed distributions (Rousselet \& Wilcox, 2020; Wilcox, 2021). The results of these analyses (see \textbf{Supplementary A}) were consistent with those of the mixed ANOVAs, which we report below. Our primary psychological outcome variables were intrinsic motivation (i.e., interest/enjoyment), perceived competence, and perceived autonomy. The mean score of the responses for these constructs at each time point was calculated for each participant and analyzed in separate mixed ANOVAs (\emph{Experiment 1:} 2 Choice x 2 Feedback x 4 Time; \emph{Experiment 2:} 2 Choice x 4 Time). Of secondary interest, error estimation abilities were assessed as total error between a participant's estimation and actual performance in pre-test (50\% of the participants in each group in Experiments 1 and 2), retention, and transfer (see \textbf{Supplementary B}).

@@

@@

Spatial (Fig. \ref{fig:fig2}A) and timing (Fig. \ref{fig:fig2}B) error decreased from the pre-test to the retention and transfer tests. There was a main effect of Test for spatial error, \(F(1.33,196.52) = 40.20\), \(p < .001\), \(\eta_{G}^2 = .138\), where performance was less errorful in retention and transfer than pre-test (\(p\)'s \textless{} .001) and performance in retention was better than transfer (\(p\) \textless{} .001). A main effect of Test was also found for timing error, \(F(1.08,160.23) = 81.21\), \(p < .001\), \(\eta_{G}^2 = .245\), with pre-test performance more errorful than both retention and transfer (\(p\)'s \textless{} .001), and retention was less errorful than transfer (\(p\) \textless{} .001). \textcolor{blue}{The main effect of Choice was not significant for both spatial, $F(1,148) = .52$, $p = .471$, $\eta_{G}^2 = .001$, and timing, $F(1,148) = .32$, $p = .547$, $\eta_{G}^2 < .001$, error.}

@@

Spatial (Fig. \ref{fig:fig3}A) and timing (Fig. \ref{fig:fig3}B) error did not change considerably from the pre-test to the retention and transfer tests. \textcolor{blue}{The main effect of Choice was not significant for both spatial, $F(1,74) = .23$, $p = .631$, $\eta_{G}^2 = .002$, and timing, $F(1,74) = .11$, $p = .738$, $\eta_{G}^2 = .001$, error.} All other main effects and interactions were also not significant.

\clearpage

@@

\caption{\small \normalfont \onehalfspacing \textbf{Experiment 1 data.} The Choice with error feedback (Choice+Error) group is shown in dark blue circles, the Choice with graded feedback (Choice+Graded) group is shown in light blue squares, the Yoked with error feedback (Yoked+Error) group is shown in red triangles, and the Yoked with graded feedback (Yoked+Graded) group is shown in yellow crosses. Error bars denote 95\% bootstrapped confidence intervals. \textbf{(A)} Spatial total error (degrees) and \textbf{(B)} timing total error (ms) averaged across blocks and participants within each group. Dotted vertical lines denote the different experimental phases. Pre-test and acquisition occurred on Day 1 and retention and transfer occurred approximately 24-hours later on Day 2. Self-reported scores for perceived autonomy \textbf{(C)}, perceived competence \textbf{(D)}, and intrinsic motivation \textbf{(E)} after the pre-test and after blocks 1 and 6 of acquisition on Day 1, and before the retention test on Day 2. Scores could range on a Likert scale from 1 (Strongly disagree) to 7 (Strongly agree). Dots represent individual data points.}\label{fig:fig2}
\end{figure}

\clearpage

@@

\caption{\small \normalfont \onehalfspacing \textbf{Experiment 2 data.} The Choice with binary feedback (Choice+Binary) group is shown in green circles and the Yoked with binary feedback (Yoked+Binary) group is shown in purple squares. Error bars denote 95\% bootstrapped confidence intervals. \textbf{(A)} Spatial total error (degrees) and \textbf{(B)} timing total error (ms) averaged across blocks and participants within each group. Dotted vertical lines denote the different experimental phases. Pre-test and acquisition occurred on Day 1 and retention and transfer occurred approximately 24-hours later on Day 2. Self-reported scores for perceived autonomy \textbf{(C)}, perceived competence \textbf{(D)}, and intrinsic motivation \textbf{(E)} after the pre-test and after blocks 1 and 6 of acquisition on Day 1, and before the retention test on Day 2. Scores could range on a Likert scale from 1 (Strongly disagree) to 7 (Strongly agree). Dots represent individual data points.}\label{fig:fig3}
\end{figure}

\clearpage
